---
title: "Optimal Tensor Transport"
author:
- name: Koki Tsuyuzaki
  affiliation: Laboratory for Bioinformatics Research,
    RIKEN Center for Biosystems Dynamics Research
  email: k.t.the-answer@hotmail.co.jp
date: "`r Sys.Date()`"
bibliography: bibliography.bib
package: guidedPLS
output: rmarkdown::html_vignette
vignette: |
  %\VignetteIndexEntry{Optimal Tensor Transport}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

# Introduction

In this vignette, we consider optimal tensor transport (OTT), which is an extension of OT to be able to handle tensors of any order by learning possibly multiple transport plans.

Here, we reproduce the experiments in the original paper [@ott]. For the details of the methodology, see the original paper.

```{r setting, echo=TRUE}
library("otTensor")

.show_matrix <- function(mat, main = ""){
    mat_rev <- apply(mat, 2, rev)
    mat_rev <- t(mat_rev)

    row_index <- 1:ncol(mat_rev)
    col_index <- 1:nrow(mat_rev)

    # grayscale
    image(mat_rev, col = gray((0:255)/255), xaxt = "n", yaxt = "n",
        xlab = "", ylab = "", axes = FALSE, main = main)
}
```

# OTT_1 (OT)
D <- 1
A <- 1
Is <- c(4)
Ks <- c(7)
f <- c(1)
arrX <- array(rep(0, prod(Is)), Is)
arrY <- array(rep(0, prod(Ks)), Ks)

for (i1 in 1:Is[1]) {
    arrX[i1] <- i1
}
for (k1 in 1:Ks[1]) {
    arrY[k1] <- k1
}

ps <- list()
for (a in 1:A) {
    ds <- which(f == a)
    d <- ds[1]
    length_of_p_a <- dim(arrX)[d]
    ps[[a]] <- rep(0.01, length_of_p_a); ps[[a]][c(1, 3)] <- 1
    ps[[a]] <- ps[[a]] / sum(ps[[a]])
}
qs <- list()
for (a in 1:A) {
    ds <- which(f == a)
    d <- ds[1]
    length_of_q_a <- dim(arrY)[d]
    qs[[a]] <- rep(1, length_of_q_a); qs[[a]][c(2, 3)] <- 0
    qs[[a]] <- qs[[a]] / sum(qs[[a]])
}

X <- as.tensor(arrX)
Y <- as.tensor(arrY)

out <- OTT(X = X, Y = Y, D = D, A = A, Is = Is, Ks = Ks, f = f,
    ps=ps, qs=qs, num.sample=1000,
    loss = function (x, y) {abs(x - y)},
    num.iter=200, epsilon=1e-10)

options(repr.plot.width=6, repr.plot.height=10)
par(mfrow=c(3, 2))
plot(ps[[1]], type="h", col="red", ylim=c(0, 1), main="ps[[1]]")
plot(qs[[1]], type="h", col="red", ylim=c(0, 1), main="qs[[1]]")
.show_matrix(out$Ts[[1]], main="Ts[[1]]")
plot(arrX, type="h", col="black", main="arrX")
plot(arrY, type="h", col="black", main="arrY")


# OTT_12 (Co-OT)

D <- 2
A <- 2
Is <- c(4, 5)
Ks <- c(7, 8)
f <- c(1, 2)
arrX <- array(rep(0, prod(Is)), Is)
arrY <- array(rep(0, prod(Ks)), Ks)

for (i1 in 1:Is[1]) {
    for (i2 in 1:Is[2]) {
        arrX[i1, i2] <- i1 + i2
    }
}
for (k1 in 1:Ks[1]) {
    for (k2 in 1:Ks[2]) {
            arrY[k1, k2] <- k1 + k2
    }
}

ps <- list()
for (a in 1:A) {
    ds <- which(f == a)
    d <- ds[1]
    length_of_p_a <- dim(arrX)[d]
    ps[[a]] <- rep(0.01, length_of_p_a); ps[[a]][c(1, 3)] <- 1
    ps[[a]] <- ps[[a]] / sum(ps[[a]])
}
qs <- list()
for (a in 1:A) {
    ds <- which(f == a)
    d <- ds[1]
    length_of_q_a <- dim(arrY)[d]
    qs[[a]] <- rep(1, length_of_q_a); qs[[a]][c(2, 3)] <- 0
    qs[[a]] <- qs[[a]] / sum(qs[[a]])
}

X <- as.tensor(arrX)
Y <- as.tensor(arrY)

out <- OTT(X = X, Y = Y, D = D, A = A, Is = Is, Ks = Ks, f = f,
    ps=ps, qs=qs, num.sample=1000,
    loss = function (x, y) {abs(x - y)},
    num.iter=200, epsilon=1e-10)

options(repr.plot.width=6, repr.plot.height=10)
par(mfrow=c(3, 2))
plot(ps[[1]], type="h", col="red", ylim=c(0, 1), main="ps[[1]]")
plot(qs[[1]], type="h", col="red", ylim=c(0, 1), main="qs[[1]]")
.show_matrix(out$Ts[[1]], main="Ts[[1]]")
.show_matrix(arrX, main="arrX")
.show_matrix(arrY, main="arrY")

par(mfrow=c(3, 2))
plot(ps[[2]], type="h", col="red", ylim=c(0, 1), main="ps[[2]]")
plot(qs[[2]], type="h", col="red", ylim=c(0, 1), main="qs[[2]]")
.show_matrix(out$Ts[[2]], main="Ts[[2]]")
.show_matrix(arrX, main="arrX")
.show_matrix(arrY, main="arrY")

# OTT_11 (GW)
D <- 2
A <- 1
Is <- c(4, 4)
Ks <- c(6, 6)
f <- c(1, 1)
arrX <- array(rep(0, prod(Is)), Is)
arrY <- array(rep(0, prod(Ks)), Ks)

for (i1 in 1:Is[1]) {
    for (i2 in 1:Is[2]) {
        arrX[i1, i2] <- i1 + i2
    }
}
for (k1 in 1:Ks[1]) {
    for (k2 in 1:Ks[2]) {
            arrY[k1, k2] <- k1 + k2
    }
}

ps <- list()
for (a in 1:A) {
    ds <- which(f == a)
    d <- ds[1]
    length_of_p_a <- dim(arrX)[d]
    ps[[a]] <- rep(0.01, length_of_p_a); ps[[a]][c(1, 3)] <- 1
    ps[[a]] <- ps[[a]] / sum(ps[[a]])
}
qs <- list()
for (a in 1:A) {
    ds <- which(f == a)
    d <- ds[1]
    length_of_q_a <- dim(arrY)[d]
    qs[[a]] <- rep(1, length_of_q_a); qs[[a]][c(2, 3)] <- 0
    qs[[a]] <- qs[[a]] / sum(qs[[a]])
}

X <- as.tensor(arrX)
Y <- as.tensor(arrY)

out <- OTT(X = X, Y = Y, D = D, A = A, Is = Is, Ks = Ks, f = f,
    ps=ps, qs=qs, num.sample=1000,
    loss = function (x, y) {abs(x - y)},
    num.iter=200, epsilon=1e-10)

options(repr.plot.width=6, repr.plot.height=10)
par(mfrow=c(3, 2))
plot(ps[[1]], type="h", col="red", ylim=c(0, 1), main="ps[[1]]")
plot(qs[[1]], type="h", col="red", ylim=c(0, 1), main="qs[[1]]")
.show_matrix(out$Ts[[1]], main="Ts[[1]]")
.show_matrix(arrX, main="arrX")
.show_matrix(arrY, main="arrY")

# OTT_111 (triplets)
D <- 3
A <- 1
Is <- c(4, 4, 4)
Ks <- c(6, 6, 6)
f <- c(1, 1, 1)
arrX <- array(rep(0, prod(Is)), Is)
arrY <- array(rep(0, prod(Ks)), Ks)

for (i1 in 1:Is[1]) {
    for (i2 in 1:Is[2]) {
        for (i3 in 1:Is[3]) {
            arrX[i1, i2, i3] <- i1 + i2 + i3
        }
    }
}
for (k1 in 1:Ks[1]) {
    for (k2 in 1:Ks[2]) {
        for (k3 in 1:Ks[3]) {
            arrY[k1, k2, k3] <- k1 + k2 + k3
        }
    }
}

ps <- list()
for (a in 1:A) {
    ds <- which(f == a)
    d <- ds[1]
    length_of_p_a <- dim(arrX)[d]
    ps[[a]] <- rep(0.01, length_of_p_a); ps[[a]][c(1, 3)] <- 1
    ps[[a]] <- ps[[a]] / sum(ps[[a]])
}
qs <- list()
for (a in 1:A) {
    ds <- which(f == a)
    d <- ds[1]
    length_of_q_a <- dim(arrY)[d]
    qs[[a]] <- rep(1, length_of_q_a); qs[[a]][c(2, 3)] <- 0
    qs[[a]] <- qs[[a]] / sum(qs[[a]])
}

X <- as.tensor(arrX)
Y <- as.tensor(arrY)

out <- OTT(X = X, Y = Y, D = D, A = A, Is = Is, Ks = Ks, f = f,
    ps=ps, qs=qs, num.sample=1000,
    loss = function (x, y) {abs(x - y)},
    num.iter=200, epsilon=1e-10)

options(repr.plot.width=6, repr.plot.height=10)
par(mfrow=c(3, 2))
plot(ps[[1]], type="h", col="red", ylim=c(0, 1), main="ps[[1]]")
plot(qs[[1]], type="h", col="red", ylim=c(0, 1), main="qs[[1]]")
.show_matrix(out$Ts[[1]], main="Ts[[1]]")
.show_matrix(arrX[,,1], main="arrX[,,1]")
.show_matrix(arrY[,,1], main="arrY[,,1]")


# OTT_123 (triCo-OT)

D <- 3
A <- 3
Is <- c(4, 5, 6)
Ks <- c(7, 8, 9)
f <- c(1, 2, 3)
arrX <- array(rep(0, prod(Is)), Is)
arrY <- array(rep(0, prod(Ks)), Ks)

for (i1 in 1:Is[1]) {
    for (i2 in 1:Is[2]) {
        for (i3 in 1:Is[3]) {
            arrX[i1, i2, i3] <- i1 + i2 + i3
        }
    }
}
for (k1 in 1:Ks[1]) {
    for (k2 in 1:Ks[2]) {
        for (k3 in 1:Ks[3]) {
            arrY[k1, k2, k3] <- k1 + k2 + k3
        }
    }
}

ps <- list()
for (a in 1:A) {
    ds <- which(f == a)
    d <- ds[1]
    length_of_p_a <- dim(arrX)[d]
    ps[[a]] <- rep(0.01, length_of_p_a); ps[[a]][c(1, 3)] <- 1
    ps[[a]] <- ps[[a]] / sum(ps[[a]])
}
qs <- list()
for (a in 1:A) {
    ds <- which(f == a)
    d <- ds[1]
    length_of_q_a <- dim(arrY)[d]
    qs[[a]] <- rep(1, length_of_q_a); qs[[a]][c(2, 3)] <- 0
    qs[[a]] <- qs[[a]] / sum(qs[[a]])
}

X <- as.tensor(arrX)
Y <- as.tensor(arrY)

out <- OTT(X = X, Y = Y, D = D, A = A, Is = Is, Ks = Ks, f = f,
    ps=ps, qs=qs, num.sample=1000,
    loss = function (x, y) {abs(x - y)},
    num.iter=200, epsilon=1e-10)

options(repr.plot.width=6, repr.plot.height=10)
par(mfrow=c(3, 2))
plot(ps[[1]], type="h", col="red", ylim=c(0, 1), main="ps[[1]]")
plot(qs[[1]], type="h", col="red", ylim=c(0, 1), main="qs[[1]]")
.show_matrix(out$Ts[[1]], main="Ts[[1]]")
.show_matrix(arrX[,,1], main="arrX[,,1]")
.show_matrix(arrY[,,1], main="arrY[,,1]")

par(mfrow=c(3, 2))
plot(ps[[2]], type="h", col="red", ylim=c(0, 1), main="ps[[2]]")
plot(qs[[2]], type="h", col="red", ylim=c(0, 1), main="qs[[2]]")
.show_matrix(out$Ts[[2]], main="Ts[[2]]")
.show_matrix(arrX[,,2], main="arrX[,,2]")
.show_matrix(arrY[,,2], main="arrY[,,2]")

par(mfrow=c(3, 2))
plot(ps[[3]], type="h", col="red", ylim=c(0, 1), main="ps[[3]]")
plot(qs[[3]], type="h", col="red", ylim=c(0, 1), main="qs[[3]]")
.show_matrix(out$Ts[[3]], main="Ts[[3]]")
.show_matrix(arrX[,,3], main="arrX[,,3]")
.show_matrix(arrY[,,3], main="arrY[,,3]")


# OTT_112 (GW Collection)
D <- 3
A <- 2
Is <- c(4, 4, 5)
Ks <- c(6, 6, 7)
f <- c(1, 1, 2)
arrX <- array(rep(0, prod(Is)), Is)
arrY <- array(rep(0, prod(Ks)), Ks)

for (i1 in 1:Is[1]) {
    for (i2 in 1:Is[2]) {
        for (i3 in 1:Is[3]) {
            arrX[i1, i2, i3] <- i1 + i2 + i3
        }
    }
}
for (k1 in 1:Ks[1]) {
    for (k2 in 1:Ks[2]) {
        for (k3 in 1:Ks[3]) {
            arrY[k1, k2, k3] <- k1 + k2 + k3
        }
    }
}

ps <- list()
for (a in 1:A) {
    ds <- which(f == a)
    d <- ds[1]
    length_of_p_a <- dim(arrX)[d]
    ps[[a]] <- rep(0.01, length_of_p_a); ps[[a]][c(1, 3)] <- 1
    ps[[a]] <- ps[[a]] / sum(ps[[a]])
}
qs <- list()
for (a in 1:A) {
    ds <- which(f == a)
    d <- ds[1]
    length_of_q_a <- dim(arrY)[d]
    qs[[a]] <- rep(1, length_of_q_a); qs[[a]][c(2, 3)] <- 0
    qs[[a]] <- qs[[a]] / sum(qs[[a]])
}

X <- as.tensor(arrX)
Y <- as.tensor(arrY)

out <- OTT(X = X, Y = Y, D = D, A = A, Is = Is, Ks = Ks, f = f,
    ps=ps, qs=qs, num.sample=1000,
    loss = function (x, y) {abs(x - y)},
    num.iter=200, epsilon=1e-10)


options(repr.plot.width=6, repr.plot.height=10)
par(mfrow=c(3, 2))
plot(ps[[1]], type="h", col="red", ylim=c(0, 1), main="ps[[1]]")
plot(qs[[1]], type="h", col="red", ylim=c(0, 1), main="qs[[1]]")
.show_matrix(out$Ts[[1]], main="Ts[[1]]")
.show_matrix(arrX[,,1], main="arrX[,,1]")
.show_matrix(arrY[,,1], main="arrY[,,1]")

par(mfrow=c(3, 2))
plot(ps[[2]], type="h", col="red", ylim=c(0, 1), main="ps[[2]]")
plot(qs[[2]], type="h", col="red", ylim=c(0, 1), main="qs[[2]]")
.show_matrix(out$Ts[[2]], main="Ts[[2]]")
.show_matrix(arrX[,,2], main="arrX[,,2]")
.show_matrix(arrY[,,2], main="arrY[,,2]")


# Session Information {.unnumbered}

```{r sessionInfo, echo=FALSE}
sessionInfo()
```

# References