---
title: "Figure2e: OTT_123 (triCo-OT)"
author:
- name: Koki Tsuyuzaki
  affiliation: Laboratory for Bioinformatics Research,
    RIKEN Center for Biosystems Dynamics Research
  email: k.t.the-answer@hotmail.co.jp
date: "`r Sys.Date()`"
bibliography: bibliography.bib
package: otTensor
output: rmarkdown::html_vignette
vignette: |
  %\VignetteIndexEntry{Figure2e: OTT_123 (triCo-OT)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

# Introduction

In this vignette, we consider optimal tensor transport (OTT), which is an extension of optimal transport (OT) to be able to handle tensors of any order by learning possibly multiple transport plans.

Original method is firstly introduced in the paper by @ott et al. For the details of the methodology, see the original paper. Here, we reproduce the Figure 2.

```{r setting, echo=TRUE}
library("otTensor")
library("rTensor")

.show_matrix <- function(mat, main = ""){
    mat_rev <- apply(mat, 2, rev)
    mat_rev <- t(mat_rev)

    row_index <- 1:ncol(mat_rev)
    col_index <- 1:nrow(mat_rev)

    # grayscale
    image(mat_rev, col = gray((0:255)/255), xaxt = "n", yaxt = "n",
        xlab = "", ylab = "", axes = FALSE, main = main)
}
```

# Figure2e: OTT_123 (triCo-OT)

Here, we use the following two tensor data.

```{r data, echo=TRUE}
D <- 3
A <- 3
Is <- c(4, 5, 6)
Ks <- c(7, 8, 9)
f <- c(1, 2, 3)
arrX <- array(rep(0, prod(Is)), Is)
arrY <- array(rep(0, prod(Ks)), Ks)

for (i1 in 1:Is[1]) {
    for (i2 in 1:Is[2]) {
        for (i3 in 1:Is[3]) {
            arrX[i1, i2, i3] <- i1 + i2 + i3
        }
    }
}
for (k1 in 1:Ks[1]) {
    for (k2 in 1:Ks[2]) {
        for (k3 in 1:Ks[3]) {
            arrY[k1, k2, k3] <- k1 + k2 + k3
        }
    }
}

ps <- list()
for (a in 1:A) {
    ds <- which(f == a)
    d <- ds[1]
    length_of_p_a <- dim(arrX)[d]
    ps[[a]] <- rep(0.01, length_of_p_a); ps[[a]][c(1, 3)] <- 1
    ps[[a]] <- ps[[a]] / sum(ps[[a]])
}
qs <- list()
for (a in 1:A) {
    ds <- which(f == a)
    d <- ds[1]
    length_of_q_a <- dim(arrY)[d]
    qs[[a]] <- rep(1, length_of_q_a); qs[[a]][c(2, 3)] <- 0
    qs[[a]] <- qs[[a]] / sum(qs[[a]])
}

X <- as.tensor(arrX)
Y <- as.tensor(arrY)
```

Then, we perform OTT by `OTT` function as follows:

```{r ott, echo=TRUE}
out <- OTT(X = X, Y = Y, f = f,
    ps=ps, qs=qs, num.sample=1000,
    loss = function (x, y) {abs(x - y)},
    num.iter=200, epsilon=1e-10)
```

The transport plans $T_1$, $T_2$, and $T_3$ estimated by `OTT` are visualized as follows:

```{r transportplan, echo=TRUE, fig.height=6, fig.width=6}
options(repr.plot.width=6, repr.plot.height=10)
oldpar <- par("mfrow")
par(mfrow=c(3, 2))
plot(ps[[1]], type="h", col="red", ylim=c(0, 1), main="ps[[1]]")
plot(qs[[1]], type="h", col="red", ylim=c(0, 1), main="qs[[1]]")
.show_matrix(out$Ts[[1]], main="Ts[[1]]")
.show_matrix(arrX[,,1], main="arrX[,,1]")
.show_matrix(arrY[,,1], main="arrY[,,1]")

par(mfrow=c(3, 2))
plot(ps[[2]], type="h", col="red", ylim=c(0, 1), main="ps[[2]]")
plot(qs[[2]], type="h", col="red", ylim=c(0, 1), main="qs[[2]]")
.show_matrix(out$Ts[[2]], main="Ts[[2]]")
.show_matrix(arrX[,,2], main="arrX[,,2]")
.show_matrix(arrY[,,2], main="arrY[,,2]")

par(mfrow=c(3, 2))
plot(ps[[3]], type="h", col="red", ylim=c(0, 1), main="ps[[3]]")
plot(qs[[3]], type="h", col="red", ylim=c(0, 1), main="qs[[3]]")
.show_matrix(out$Ts[[3]], main="Ts[[3]]")
.show_matrix(arrX[,,3], main="arrX[,,3]")
.show_matrix(arrY[,,3], main="arrY[,,3]")
par(mfrow=oldpar)
```

# Session Information {.unnumbered}

```{r sessionInfo, echo=FALSE}
sessionInfo()
```

# References